---
title: "p8105_hw3_sc5154"
output: github_document
date: '2022-10-15'
---

```{r}
library(tidyverse)
```

# Problem 2

## Question 1

> Load, tidy, and otherwise wrangle the data. Your final dataset should include all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes. Describe the resulting dataset (e.g. what variables exist, how many observations, etc).

```{r}
accel_df = read_csv("data/accel_data.csv") %>%
  janitor::clean_names() %>%
  pivot_longer(
    activity_1:activity_1440,
    names_to = 'activity_index',
    names_prefix = 'activity_',
    values_to = 'activity_count'
  ) %>%
  mutate(
  day = factor(day, levels = c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday')),
  is_weekday = ifelse((day == 'Saturday' | day == 'Sunday'), 0, 1),
  activity_index = as.numeric(activity_index)
  ) %>%
  arrange(week, day_id)
head(accel_df)
```

The final dataset contains `r nrow(accel_df)` rows (observations) and `r ncol(accel_df)` columns. The variables are `r variable.names(accel_df)`. 'is_weekday' indicates weekday vs. weekend. 'activity_index' and 'activity_count' are obtained by transforming from the original dataset for better clarity.

## Qustion 2

> Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?

```{r}
accel_df %>%
  group_by(week, day) %>%
  summarize(daily_activity_total = sum(activity_count)) %>%
  arrange(week, day) %>%
  pivot_wider(
    names_from = day,
    values_from = daily_activity_total
  ) %>%
  knitr::kable()
```

Note that the daily total activity counts on Saturday for week 4 and 5 are both 1440, which is equal to the number of minutes in a day. This seems abnormal and should be considered outliers. Apart from that, no obvious trend or pattern is observed from the table.

## Question 3

> Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.

```{r}
accel_df %>%
  mutate(
    daily_time = floor((activity_index)/60) +
      (1/60) * (activity_index) %% 60
    ) %>%
  ggplot(aes(x = daily_time, y = activity_count, color = day)) +
    geom_line(alpha = 0.3) + 
    theme(legend.position = 'bottom') + 
    scale_x_continuous(limits = c(0,24), breaks = seq(0, 24, 2)) +
    labs(x = 'Time', 
         y = 'Activity Counts', 
         color = 'Day of the Week')
```

Activity counts are minimum between 12 am and 4 am and are maximum between 8 pm and 10 pm. They is also a local maximum between 10 am and 12 pm.

# Problem 3

## Question 1

> Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?

## Question 2

> Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?

## Question 3

> Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

