---
title: "p8105_hw3_sc5154"
output: github_document
date: '2022-10-15'
---

```{r}
library(p8105.datasets)
library(tidyverse)
library(ggridges)
library(patchwork)
```

# Problem 2

## Question 1

> Load, tidy, and otherwise wrangle the data. Your final dataset should include all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes. Describe the resulting dataset (e.g. what variables exist, how many observations, etc).

```{r}
accel_df = read_csv("data/accel_data.csv") %>%
  janitor::clean_names() %>%
  pivot_longer(
    activity_1:activity_1440,
    names_to = 'activity_index',
    names_prefix = 'activity_',
    values_to = 'activity_count'
  ) %>%
  mutate(
  day = factor(day, levels = c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday')),
  is_weekday = ifelse((day == 'Saturday' | day == 'Sunday'), 0, 1),
  activity_index = as.numeric(activity_index)
  ) %>%
  arrange(week, day_id)
head(accel_df)
```

The final dataset contains `r nrow(accel_df)` rows (observations) and `r ncol(accel_df)` columns. The variables are `r variable.names(accel_df)`. 'is_weekday' indicates weekday vs. weekend. 'activity_index' and 'activity_count' are obtained by transforming from the original dataset for better clarity.

## Qustion 2

> Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?

```{r}
accel_df %>%
  group_by(week, day) %>%
  summarize(daily_activity_total = sum(activity_count)) %>%
  arrange(week, day) %>%
  pivot_wider(
    names_from = day,
    values_from = daily_activity_total
  ) %>%
  knitr::kable()
```

Note that the daily total activity counts on Saturday for week 4 and 5 are both 1440, which is equal to the number of minutes in a day. This seems abnormal and should be considered outliers. Apart from that, no obvious trend or pattern is observed from the table.

## Question 3

> Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.

```{r}
accel_df %>%
  mutate(
    daily_time = floor((activity_index)/60) +
      (1/60) * (activity_index) %% 60
    ) %>%
  ggplot(aes(x = daily_time, y = activity_count, color = day)) +
    geom_line(alpha = 0.3) + 
    theme(legend.position = 'bottom') + 
    scale_x_continuous(limits = c(0,24), breaks = seq(0, 24, 2)) +
    labs(x = 'Time', 
         y = 'Activity Counts', 
         color = 'Day of the Week')
```

Activity counts are minimum between 12 am and 4 am and are maximum between 8 pm and 10 pm. They is also a local maximum between 10 am and 12 pm.

# Problem 3

## Question 0

> Write a short description of the dataset, noting the size and structure of the data, describing some key variables, and indicating the extent to which missing data is an issue.

```{r}
data("ny_noaa")
```

The original dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Variables are `r variable.names(ny_noaa)`.

## Question 1

> Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?

```{r}
noaa_df = 
  ny_noaa %>%
  janitor::clean_names() %>%
  separate(date, c("year", "month", "day"), sep = "-", convert = TRUE) %>%
  mutate(
    month = month.name[as.integer(month)],
    prcp = prcp / 10,
    tmax = as.numeric(tmax) / 10, 
    tmin = as.numeric(tmin) / 10
    )
```

The final dataset contains `r nrow(noaa_df)` rows and `r ncol(noaa_df)` columns. Variables are `r variable.names(ny_noaa)`.

'prcp', 'snow', 'snwd', 'tmax' and 'tmin' indicates precipitation (tenths of mm), snowfall (mm), snow depth (mm), maximum temperature (tenths of degrees C), and minimum temperature (tenths of degrees C), respectively.

```{r}
most_snow = 
  noaa_df %>%
  group_by(snow) %>%
  summarize(n_obs = n()) %>%
  arrange(-n_obs) %>%
  head(1)
```

The most commonly observed values for snowfall are `r most_snow$snow` with `r most_snow$n_obs` observations because NY doesn't often snow.

## Question 2

> Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?

```{r}
noaa_df %>%
  filter(month == 'January' | month == 'July') %>%
  group_by(month, id, year) %>%
  summarize(mean_tmax = mean(tmax, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = mean_tmax, color = id)) +
  geom_point(size = 0.2, alpha = 0.5) +
  geom_path(alpha = 0.5) +
  theme(legend.position = "none") + 
  labs(x = 'Year', y = "Mean Max Temperature") +
  facet_grid(. ~ month)
```

The mean max temperature in January range from -10 to 10 degree C, while in July it ranges from 20 to 35 degree C. Therefore, the mean max temperature is generally higher in July compared to January.

Note there are some outliers in January, for exmaple, the extremely low value below -13. There are also outliers in July, for example, the exteremy low value below 15.

## Question 3

> Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

```{r}
tmax_vs_tmin = 
  noaa_df %>%
  ggplot(aes(x = tmax, y = tmin)) +
  geom_hex() + 
  labs(x = 'Max Temperature', y = "Min Temperature")
snowfall_dis = 
  noaa_df %>%
  filter(snow > 0 & snow < 100) %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(x = snow, y = year)) + 
  geom_density_ridges(scale = 0.8) +
  labs(x = 'Snowfall', y = "Year")
tmax_vs_tmin + snowfall_dis
```